Clines from PDE

### 2015-11-22 notes/pde-clines.Rmd
solving haplotypes

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(ReacTran)
library(Matrix)
source("tran.1D.R") # fix for log.A
```

@@

Writing $g_A$ and $g_B$ for the moment, and dropping $x$ and $\tau$,
the generator of the recombination part is, 
where $x\wedge y=\max(x,y)$ and $x \vee y = \min(x,y)$,
$$\begin{aligned}
Ig_A(a,b)
    &= 
        p \int_{a\wedge 0}^b g_A(a,\theta) g_A(\theta,b) d\theta
        + (1-p) \int_{a\wedge 0}^b g_A(a,\theta) g_B(\theta,b) d\theta
        + p \int_{a}^{b \vee 0} g_A(a,\theta) g_A(\theta,b) d\theta
        + (1-p) \int_{a\wedge 0}^b g_B(a,\theta) g_A(\theta,b) d\theta 
       \\ &\qquad {}
         -(b-a)g_A(a,b) \\
    &= 
    \int_a^b
        p \one_{\theta>0} g_A(a,\theta) g_A(\theta,b) 
        + (1-p) \one_{\theta>0} g_A(a,\theta) g_B(\theta,b) 
        + p \one_{\theta\le0} g_A(a,\theta) g_A(\theta,b) 
        + (1-p) \one_{\theta\le0} g_B(a,\theta) g_A(\theta,b) d\theta
       \\ &\qquad {}
         -(b-a)g_A(a,b) \\
    &= 
    \int_a^b
    \left\{
        p g_A(a,\theta) g_A(\theta,b) 
        + (1-p) \left( 
            \one_{\theta>0} g_A(a,\theta) g_B(\theta,b) 
            + \one_{\theta\le0} g_B(a,\theta) g_A(\theta,b) 
        \right) 
        -g_A(a,b) \right\}
        d\theta   \\
    &= 
    p \int_a^b
        \left( g_A(a,\theta) g_A(\theta,b)  - g_A(a,b) \right) d\theta 
    \\ &\qquad {}
    + (1-p) \int_a^b \one_{\theta>0} \left( g_A(a,\theta) g_B(\theta,b) - g_A(a,b) \right) d\theta
    \\ &\qquad {}
    + (1-p) \int_a^b \one_{\theta\le0} \left( g_B(a,\theta) g_A(\theta,b) - g_A(a,b) \right) d\theta  .
\end{aligned}$$
$Ig_B$ is the same after exchanging $A$ with $B$, and $p$ with $1-p$.

@@

Note that if we think of $u(a,b)$ as a kernel, 
then the map $u(a,b) \mapsto \int_a^b u(a,\theta) u(\theta,b) d\theta$
is similar to the product of the kernel to itself;
on discretization this becomes like a matrix power.
This is implemented in some C code.
<!--
```{r solve_branching}
source("unifprod.R")  # for the C part
recomb_generator <- function (x,eps) {
    # the discretization of the map y(a,b) -> \int_a^b (h(a,u)*h(u,b)-h(a,b)) du
    # x is the upper triangular part of a square matrix
    #   eps is the grid spacing between locations
    x2 <- eps*unifprod_ut(x)
}
xgrid <- setup.grid.1D(x.up=-20, x.down=20, N=10)
agrid <- setup.grid.1D(x.up=0.0, x.down=0.5, N=5)
# function to find a,b coordinates from the upper triangular coordinates
ab.ord <- matrix( 0, nrow=agrid$N, ncol=agrid$N )
avals <- agrid$x.mid[row(ab.ord)[upper.tri(ab.ord,diag=TRUE)]]
bvals <- agrid$x.mid[col(ab.ord)[upper.tri(ab.ord,diag=TRUE)]]
rvals <- ifelse( avals*bvals<0, 0, pmin(avals,bvals) )
ab.coord <- function (ab) { cbind( a=avals[ab], b=bvals[ab] ) }
# we'll save the solution as an array,
#  with dimensions (x,ab)
# Here ab is upper-triangular ordered

s <- 0.1
tt <- seq(0,3,length.out=5)
fwds.soln <- forwards_pde(s=s,times=tt,grid=xgrid)

soln.array <- matrix(0.0,nrow=2*xgrid$N,ncol=agrid$N*(agrid$N+1)/2)

    sigma=1
    yinit=rep( c( rep(1.0,xgrid$N), rep(0.0,xgrid$N) ), ncol(soln.array) )

    eps <- 1e-16
    rev.pde.fn <- function (t,y,parms,...) {
        # need to apply tran.1D to columns
        # and recomb_generator to rows
        dim(y) <- dim(soln.array)
        p <- cline_interp(t,soln=fwds.soln) # freq of B
        log.p <- log(pmax(p,min(eps,min(p[p>0]))))
        log.1mp <- log(pmax(1-p,min(eps,min((1-p)[p<1]))))
        diffusion <- sapply( 1:ncol(y), function (kcol) {
                r <- rvals[kcol]
                yA <- y[1:xgrid$N,kcol]
                yB <- y[xgrid$N+(1:xgrid$N),kcol]
                tran.A <- tran.1D(C=yA, A=log.p, D=sigma^2/2, dx=xgrid, flux.up=0, flux.down=0, log.A=TRUE)$dC
                tran.B <- tran.1D(C=yB, A=log.1mp, D=sigma^2/2, dx=xgrid, flux.up=0, flux.down=0, log.A=TRUE)$dC
                c( 
                        tran.A + r * (1-p) * (yB-yA),
                        tran.B + r * p * (yA-yB)
                        )
             } )
        recombination <- apply( y, 1, recomb_generator, p=p )
        return( list( diffusion + recombination ) )
    }
    rev.soln <- ode.1D( y=yinit, times=times, func=rev.pde.fn, nspec=2*ncol(soln.array), tcrit=max(fwds.soln[,1]) )

# determines the order the solution is in:
xab <- subset( expand.grid( x=xgrid$x.mid, a=agrid$x.mid, b=agrid$x.mid ), ( b>0 ) & (abs(a)<=abs(b)) )
find_ab <- function (ab) {
    # map into the subset of (a,b) contained in xab
    flip <- ab[,1]>ab[,2] 
    ab[flip,] <- ab[flip,] %*% cbind( c(0,1), c(1,0) )
    flop <- ( ab[,1]<0 & abs(ab[,1])>=abs(ab[,2]) )
    ab[flop,] <- ab[flop,] %*% cbind( c(0,-1), c(-1,0) )
    return(ab)
}
```

-->

>>>>

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
library(ReacTran)
library(Matrix)
library(parallel)
source("tran.1D.R") # fix for log.A
```

@@

Writing $g_A$ and $g_B$ for the moment, and dropping $x$ and $\tau$,
the generator of the recombination part is, 
where $x\wedge y=\max(x,y)$ and $x \vee y = \min(x,y)$,
$$\begin{aligned}
Ig_A(a,b)
    &= 
        p \int_{a\wedge 0}^b g_A(a,\theta) g_A(\theta,b) d\theta
        + (1-p) \int_{a\wedge 0}^b g_A(a,\theta) g_B(\theta,b) d\theta
        + p \int_{a}^{b \vee 0} g_A(a,\theta) g_A(\theta,b) d\theta
        + (1-p) \int_{a\wedge 0}^b g_B(a,\theta) g_A(\theta,b) d\theta 
       \\ &\qquad {}
         -(b-a)g_A(a,b) \\
    &= 
    \int_a^b
        p \one_{\theta>0} g_A(a,\theta) g_A(\theta,b) 
        + (1-p) \one_{\theta>0} g_A(a,\theta) g_B(\theta,b) 
        + p \one_{\theta\le0} g_A(a,\theta) g_A(\theta,b) 
        + (1-p) \one_{\theta\le0} g_B(a,\theta) g_A(\theta,b) d\theta
       \\ &\qquad {}
         -(b-a)g_A(a,b) \\
    &= 
    \int_a^b
    \left\{
        p g_A(a,\theta) g_A(\theta,b) 
        + (1-p) \left( 
            \one_{\theta>0} g_A(a,\theta) g_B(\theta,b) 
            + \one_{\theta\le0} g_B(a,\theta) g_A(\theta,b) 
        \right) 
        -g_A(a,b) \right\}
        d\theta   
\end{aligned}$$
$Ig_B$ is the same after exchanging $A$ with $B$, and $p$ with $1-p$.

@@

Note that if we think of $u(a,b)$ as a kernel, 
then the map $u(a,b) \mapsto \int_a^b u(a,\theta) u(\theta,b) d\theta$
is similar to the product of the kernel to itself;
on discretization this becomes like a matrix power.
This is implemented in some C code.
```{r solve_branching, cache=TRUE}
recomb_generator

s <- 0.1
tt <- seq(0,10,length.out=20)

xgrid <- setup.grid.1D(x.up=-20, x.down=20, N=10)
rgrid <- setup.grid.1D(x.up=-0.5, x.down=0.5, N=50)

# took 24min with xgrid$N=10, rgrid$N=20 and 50 time steps up to t=3, :(
hap.soln <- forwards_backwards_haplotypes(s=s, times=tt, xgrid=xgrid, rgrid=rgrid)
```

```{r plot_hap_probs, cache=TRUE, depends="solve_branching"}
hap.probs <- get_hap_probs( hap.soln, left=-.025, right=.025 )
image( hap.probs, zlim=c(0,1), legend=TRUE, yaxt='n', 
      main=paste(c("A:","B:"),paste(round(attr(hap.probs,"r"),3),collapse=" to ")) )
axis(2,at=(pretty(xgrid$x.mid)-min(xgrid$x.mid))/diff(range(xgrid$x.mid)),labels=pretty(xgrid$x.mid))
```
